---
title: "Soil biogeochemistry"
date: '2018-01-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE)
setwd("")
```

### Load and subset the data

Load the data set 'biogeochemistry.csv', which contains the following biogeochemical measurements taken at the agroforestry site, split into crop and tree lines, and the forest site, split into upper and lower slopes:

- pH
- Soil organic C
- Bulk density
- Water-filled pore space
- Total N
- Total C:N ratio
- Microbial biomass C
- Microbial biomass N
- Microbial biomass C:N ratio
- NH4+
- NO3-
- K2SO4-
- N fixing rate

```{r}
d <- read.csv("data/biogeochemistry.csv", header = TRUE, sep = "\t", dec = ",")

agro <- subset(d, sites=="CA_C" | sites=="CA_T")
crop <- subset(agro, sites=="CA_C")
tree <- subset(agro, sites=="CA_T")
forest <- subset(d, sites=="F_U" | sites=="F_L")
upper <- subset(forest, sites=="F_U")
lower <- subset(forest, sites=="F_L")
```

### Comparison means of two groups

First, we compare the differences in various biogeochemical metrics between the crop and tree lines in the agroforestry site and the upper and lower lines in the forest site.

We would check that what assumptions for statistical tests are met, namely equal variances and the normal distribution.

A visual inspection using histograms or quantile-quantile plots as well as Shapiro's test can be used to see if normal distribution is likely.

Note that small sample sizes make it more difficult and less sensible to judge whether the data is normally distributed.

When in doubt, use non-parametric tests with fewer assumptions like the Wilcoxson test.

```{r pH}
var.test(crop$pH, tree$pH) 
shapiro.test(agro$pH)
t.test(crop$pH, tree$pH, var.equal = TRUE)

var.test(upper$pH, lower$pH) 
shapiro.test(forest$pH)
t.test(upper$pH, lower$pH, var.equal = TRUE)
```

```{r SOC}
var.test(crop$SOC, tree$SOC)
shapiro.test(agro$SOC)
t.test(crop$SOC, tree$SOC, var.equal = TRUE)

var.test(upper$SOC, lower$SOC)
shapiro.test(forest$SOC)
t.test(upper$SOC, lower$SOC, var.equal = TRUE)
```

```{r BD}
var.test(crop$BD, tree$BD)
shapiro.test(agro$BD)
t.test(crop$BD, tree$BD, var.equal = TRUE)

var.test(upper$BD, lower$BD)
shapiro.test(forest$BD)
wilcox.test(upper$BD, lower$BD)
```

```{r WFPS}
var.test(crop$WFPS, tree$WFPS)
shapiro.test(agro$WFPS)
t.test(crop$WFPS, tree$WFPS, var.equal = TRUE)

var.test(upper$WFPS, lower$WFPS)
shapiro.test(forest$WFPS)
t.test(upper$WFPS, lower$WFPS)
```

```{r N}
var.test(crop$N, tree$N)
shapiro.test(agro$N)
wilcox.test(crop$N, tree$N)

var.test(upper$N, lower$N)
shapiro.test(forest$N)
t.test(upper$N, lower$N, var.equal = TRUE)
```

```{r CN_ratio}
var.test(crop$CN_ratio, tree$CN_ratio)
shapiro.test(agro$CN_ratio)
t.test(crop$CN_ratio, tree$CN_ratio, var.equal = TRUE)

var.test(upper$CN_ratio, lower$CN_ratio)
shapiro.test(forest$CN_ratio)
t.test(upper$CN_ratio, lower$CN_ratio, var.equal = TRUE)
```

```{r MB_C}
var.test(crop$MB_C, tree$MB_C)
shapiro.test(agro$MB_C)
t.test(crop$MB_C, tree$MB_C, var.equal = TRUE)

var.test(upper$MB_C, lower$MB_C)
shapiro.test(forest$MB_C)
t.test(upper$MB_C, lower$MB_C, var.equal = TRUE)
```

```{r MB_N}
var.test(crop$MB_N, tree$MB_N)
shapiro.test(agro$MB_N)
t.test(crop$MB_N, tree$MB_N, var.equal = TRUE)

var.test(upper$MB_N, lower$MB_N)
shapiro.test(forest$MB_N)
t.test(upper$MB_N, lower$MB_N, var.equal = TRUE)
```

```{r MB_CN}
var.test(crop$MB_CN, tree$MB_CN)
shapiro.test(agro$MB_CN)
t.test(crop$MB_CN, tree$MB_CN, var.equal = TRUE)

var.test(upper$MB_CN, lower$MB_CN)
shapiro.test(forest$MB_CN)
t.test(upper$MB_CN, lower$MB_CN, var.equal = TRUE)
```

```{r NH4}
var.test(crop$NH4, tree$NH4)
shapiro.test(agro$NH4)
wilcox.test(crop$NH4, tree$NH4)

var.test(upper$NH4, lower$NH4)
shapiro.test(forest$NH4)
wilcox.test(upper$NH4, lower$NH4)
```

```{r NO3}
var.test(crop$NO3, tree$NO3)
shapiro.test(agro$NO3)
t.test(crop$NO3, tree$NO3, var.equal = TRUE)

var.test(upper$NO3, lower$NO3)
t.test(crop$NO3, tree$NO3, var.equal = FALSE)
```

```{r K2SO4}
var.test(crop$K2SO4, tree$K2SO4)
shapiro.test(agro$K2SO4)
wilcox.test(crop$K2SO4, tree$K2SO4)

var.test(upper$K2SO4, lower$K2SO4)
shapiro.test(forest$K2SO4)
t.test(upper$K2SO4, lower$K2SO4, var.equal = TRUE)
```

```{r N_fix_rate}
var.test(crop$N_fix_rate, tree$N_fix_rate)
t.test(crop$N_fix_rate, tree$N_fix_rate, var.equal = FALSE)

var.test(upper$N_fix_rate, lower$N_fix_rate)
shapiro.test(forest$N_fix_rate)
wilcox.test(upper$N_fix_rate, lower$N_fix_rate)
```

The following biogeochemical measures are significantly different between the crop and tree lines at the agroforestry site:

- Water-filled pore space
- Total C:N ratio
- NO3-
- N fixing rate

The following biogeochemical measures are significantly different between the upper and lower slopes at the forest site:

- Soil organic C
- Bulk density
- NO3-

### Comparison means of more than two groups

Next, we compare the differences in various biogeochemical metrics between the crop and tree lines in the agroforestry site and the upper and lower lines in the forest site.

We would check that what assumptions for statistical tests are met, namely equal variances and the normal distribution of residuals.

Depending on which assumptions are met, we will run the ANOVA, the one-way ANOVA or the Kruskal-Wallis test.

If the null hypothesis is rejected, we will use respective post-hoc tests to see where the difference is.

```{r}
library(car)
library(pgirmess)
```

```{r pH}
leveneTest(d$pH~d$sites)
shapiro.test(residuals(lm(d$pH~d$sites)))

model <- aov(d$pH~d$sites)
anova(model)

TukeyHSD(model)
```

```{r SOC}
leveneTest(d$SOC~d$sites)
shapiro.test(residuals(lm(d$SOC~d$sites)))

model <- aov(d$SOC~d$sites)
anova(model)

TukeyHSD(model)
```

```{r BD}
leveneTest(d$BD~d$sites)
shapiro.test(residuals(lm(d$BD~d$sites)))

model <- aov(d$BD~d$sites)
anova(model)

TukeyHSD(model)
```

```{r WFPS}
leveneTest(d$WFPS~d$sites)
shapiro.test(residuals(lm(d$WFPS~d$sites)))

model <- aov(d$WFPS~d$sites)
anova(model)

TukeyHSD(model)
```

```{r N}
leveneTest(d$N~d$sites)
shapiro.test(residuals(lm(d$N~d$sites)))

model <- aov(d$N~d$sites)
anova(model)
```

```{r CN}
leveneTest(d$CN~d$sites)
shapiro.test(residuals(lm(d$CN~d$sites)))

model <- aov(d$CN~d$sites)
anova(model)

TukeyHSD(model)
```

```{r MB_C}
leveneTest(d$MB_C~d$sites)
shapiro.test(residuals(lm(d$MB_C~d$sites)))

model <- aov(d$MB_C~d$sites)
anova(model)
```

```{r MB_N}
leveneTest(d$MB_N~d$sites)
shapiro.test(residuals(lm(d$MB_N~d$sites)))

model <- aov(d$MB_N~d$sites)
anova(model)
```

```{r MB_CN}
leveneTest(d$MB_CN~d$sites)
shapiro.test(residuals(lm(d$MB_CN~d$sites)))

model <- aov(d$MB_CN~d$sites)
anova(model)

TukeyHSD(model)
```

```{r NH4}
leveneTest(d$NH4~d$sites)
shapiro.test(residuals(lm(d$NH4~d$sites)))

kruskal.test(d$NH4~d$sites)
```

```{r NO3}
leveneTest(d$NO3~d$sites)
shapiro.test(residuals(lm(d$NO3~d$sites)))

kruskal.test(d$NO3~d$sites)

kruskalmc(d$NO3~d$sites, probs=0.05, cont=NULL)
```

```{r K2SO4}
leveneTest(d$K2SO4~d$sites)
shapiro.test(residuals(lm(d$K2SO4~d$sites)))

kruskal.test(d$K2SO4~d$sites)

kruskalmc(d$K2SO4~d$sites, probs=0.05, cont=NULL)
```

```{r N_fix_rate}
leveneTest(d$N_fix_rate~d$sites)

oneway.test(d$N_fix_rate~d$sites, var.equal = FALSE)

pairwise.t.test(d$N_fix_rate, d$sites, pool.sd = FALSE, paired = FALSE, p.adjust.method = "holm")
```

The following biogeochemical measurements are significantly different between the agroforestry and forest sites:

- pH
- Soil organic C
- Bulk density
- Total C:N ratio
- Microbial biomass C:N ratio

The following biogeochemical measurement is significantly different between the crop lines in the agroforestry site and the forest site:

- Water-filled pore space

The following biogeochemical measurement is significantly different between the crop lines in the agroforestry site and the lower slopes of the forest site:

- NO3-

The following biogeochemical measurement is significantly different between the tree lines in the agroforestry site and the lower slopes of the forest site:

- K2SO4-

There is no significant differences between the agroforestry and forest sites:

- Total N
- Microbial biomass C
- Microbial biomass N
- NH4+
- N fixing rate
